---
title: "Radix trees in R"
author: "Oliver Keyes"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Radix trees in R}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

A **radix tree**, or **trie**, is a data structure optimised for storing key-value pairs in a way optimised for searching. This makes them very, very good for efficiently matching data against keys, and retrieving the values *associated* with those keys.

`triebeard` provides an implementation of tries for R (and one that can be used in Rcpp development, too, if that's your thing) so that useRs can take advantage of the fast, efficient and user-friendly matching that they allow.

## Radix usage

Suppose we have observations in a dataset that are labelled, with a 2-3 letter code that identifies the facility the sample came from:

```{r, eval=FALSE}
labels <- c("AO-1002", "AEO-1004", "AAI-1009", "AFT-1403", "QZ-9065", "QZ-1021", "RF-0901",
            "AO-1099", "AFT-1101", "QZ-4933")
```

We know the facility each code maps to, and we want to be able to map the labels to that - not over 10 entries but over hundreds, or thousands, or hundreds *of* thousands. Tries are a great way of doing that: we treat the codes as *keys* and the full facility names as *values*. So let's make a trie to do this matching, and then, well, match:

```{r, eval=FALSE}
library(triebeard)
trie <- create_trie(keys = c("AO", "AEO", "AAI", "AFT", "QZ", "RF"),
                    values = c("Audobon", "Atlanta", "Ann Arbor", "Austin", "Queensland", "Raleigh"))

longest_match(trie = trie, to_match = labels)

 [1] "Audobon"    "Atlanta"    "Ann Arbor"  "Austin"     "Queensland" "Queensland" "Raleigh"    "Audobon"    "Austin"    
[10] "Queensland"
````

This pulls out, for each label, the trie value where the associated key has the longest prefix-match to the label. We can also just grab all the values where the key starts with, say, A:

```{r, eval=FALSE}
prefix_match(trie = trie, to_match = "A")

[[1]]
[1] "Ann Arbor" "Atlanta"   "Austin"    "Audobon"  
```

And finally if we want we can match very, very fuzzily using "greedy" matching:

```{r, eval=FALSE}
greedy_match(trie = trie, to_match = "AO")

[[1]]
[1] "Ann Arbor" "Atlanta"   "Austin"    "Audobon"  
```

These operations are very, very efficient. If we use longest-match as an example, since that's the most useful thing, with a one-million element vector of things to match against:

```{r, eval=FALSE}
library(triebeard)
library(microbenchmark)

trie <- create_trie(keys = c("AO", "AEO", "AAI", "AFT", "QZ", "RF"),
                    values = c("Audobon", "Atlanta", "Ann Arbor", "Austin", "Queensland", "Raleigh"))

labels <- rep(c("AO-1002", "AEO-1004", "AAI-1009", "AFT-1403", "QZ-9065", "QZ-1021", "RF-0901",
                "AO-1099", "AFT-1101", "QZ-4933"), 100000)

microbenchmark({longest_match(trie = trie, to_match = labels)})

Unit: milliseconds
                                                  expr      min       lq     mean   median       uq      max neval
 {     longest_match(trie = trie, to_match = labels) } 284.6457 285.5902 289.5342 286.8775 288.4564 327.3878   100
```

I think we can call <300 milliseconds for a million matches against an entire set of possible values pretty fast.

## Radix modification and metadata

There's always the possibility that (horror of horrors) you'll have to add or remove entries from the trie. Fear not; you can do just that with `trie_add` and `trie_remove` respectively, both of which silently modify the trie they're provided with to add or remove whatever key-value pairs you provide.

You can also extract information from tries without using them, including the size (through `length`), the keys, or the values, and we can demonstrate all of those things nicely here:

```{r, eval=FALSE}
trie <- create_trie(keys = c("AO", "AEO", "AAI", "AFT", "QZ", "RF"),
                    values = c("Audobon", "Atlanta", "Ann Arbor", "Austin", "Queensland", "Raleigh"))
length(trie)
[1] 6

get_keys(trie)
[1] "AAI" "AEO" "AFT" "AO"  "QZ"  "RF" 

get_values(trie)
[1] "Ann Arbor"  "Atlanta"    "Austin"     "Audobon"    "Queensland" "Raleigh" 

trie_add(trie, "BT", "Breton")
length(trie)

[1] 7

trie_remove(trie, "BT")
length(trie)

[1] 6
```


### Other trie things

If you have ideas for other trie-like structures, or functions that would be useful with *these* tries, the best approach
is to either [request it](https://github.com/Ironholds/triebeard/issues) or [add it](https://github.com/Ironholds/triebeard/pulls)!
